---
title: "R Notebook"
output: html_notebook
---


### Exploratory Data Analysis: 2011 Data

```{r,echo=FALSE}
length(data_2011$Regular.pay) # total number of people in the department
data.frame(table(data_2011$Job.title)) # number of people with each title
```

```{r, echo=FALSE}
data_2011 %>%
  group_by(Job.title) %>%
  summarize(min(Regular.pay), mean(Regular.pay), max(Regular.pay))
```

```{r, echo=FALSE}
ggplot(aes(x = Regular.pay), data = data_2011) +
  geom_histogram(bins = 30)
```

```{r, echo=FALSE}
ggplot(aes(x = Regular.pay), data = data_2011) +
  geom_histogram() + 
  facet_wrap(~ Job.title)
```

```{r, echo=FALSE}
# Who is the person that made no money and how were they compensated otherwise?
ggplot(aes(x = Regular.pay), data = subset(data_2011, Job.title == "AUDITOR-APPRAISER I")) +
  geom_histogram(bins = 30)
```

### Observations for Regular Pay:

The total number of people working for the appraiser's department in 2011 was 153.
Employees with the title of appraiser fall into 12 categories. The mean regular
salary for all employees with the title appraiser is $72680.  One employee, with
the title Auditor-Appraiser I was given no Regular Pay.  From the distribution of the 
Auditor-Appraiser I Regular Pay below, it seems like this entry was a clerical error.
Additionally, the Principle Audior-Appraiser has the lowest salary with around $709.84.


```{r, echo=FALSE}
# Total Benefits

ggplot(aes(x=Total.benefits), data=data_2011) +
  geom_histogram(bins=30)
```

```{r, echo=FALSE}
data_2011 %>%
  group_by(Job.title) %>%
  summarize(min(Total.benefits), mean(Total.benefits), max(Total.benefits))
```

```{r, echo=FALSE}
# Grouping the Total Pay
grouped_pay <- total_data %>%
  group_by(Job.title, Year) %>%
  summarize(mean(Total.pay), length(Name), mean(Total.pay)/length(Name))

# Grouping the Total Benefits
grouped_benefits <- total_data %>%
  group_by(Job.title, Year) %>%
  summarize(mean(Total.benefits), length(Name), mean(Total.benefits)/length(Name))

# Writing in better column names
colnames(grouped_pay) <- c("Job.title", "Year", "Mean.pay", "Num.people", "Mean.pay.per.person")
colnames(grouped_benefits) <- c("Job.title", "Year", "Mean.benefits", "Num.people", "Mean.benefits.per.person")
```