
## Orange County Appraiser Salary Analysis

Author: Brenden Prieto

This notebook explores the trends in salary for employees
of Orange County's Appraiser's Office from 2011 until 2016.
The data was extracted from transparentcalifornia.com.

Two statistics are investigated in this report:

1. How has employee compensation for each role changed over the 5 year period? How has real salary changed relative to benefits.
2. How long does it take for appraiser's at each level to promote to the next level.

In the future, I'd like to examine the total county budget for each of these years and 
how the demographics of the county appraiser's office has changed.  How many people
who started in 2011 stayed until 2016? Male vs Female split (this would take a
while to find because I would need to code the gender of each employee).

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(RColorBrewer)
```

```{r, echo=FALSE}
# Reading in the data
data_2011 <- read.csv(file="./data/appraiser_salaries_2011.csv", header=TRUE, sep=",")
data_2012 <- read.csv(file="./data/appraiser_salaries_2012.csv", header=TRUE, sep=",")
data_2013 <- read.csv(file="./data/appraiser_salaries_2013.csv", header=TRUE, sep=",")
data_2014 <- read.csv(file="./data/appraiser_salaries_2014.csv", header=TRUE, sep=",")
data_2015 <- read.csv(file="./data/appraiser_salaries_2015.csv", header=TRUE, sep=",")
data_2016 <- read.csv(file="./data/appraiser_salaries_2016.csv", header=TRUE, sep=",")
```

```{r, echo=FALSE}
# Helper function for cleaning the datasets
clean_data <- function(df, year) {
  df <- df[df$Name != "",] # Deleting empty rows
  df$Regular.pay <- as.numeric(gsub("\\$|,", "", df$Regular.pay))
  df$Overtime.pay <- as.numeric(gsub("\\$|,", "", df$Overtime.pay))
  df$Other.pay <- as.numeric(gsub("\\$|,", "", df$Other.pay))
  df$Total.pay <- as.numeric(gsub("\\$|,", "", df$Total.pay))
  df$Total.benefits <- as.numeric(gsub("\\$|,", "", df$Total.benefits))
  df$Total.pay...benefits <- as.numeric(gsub("\\$|,", "", df$Total.pay...benefits))
  
  # Cleaning the Job titles
  df$Job.title <- toupper(gsub(paste("\n\t\t\t\t\tOrange County,", year, sep=' '), "", df$Job.title))
  
  # Adding a year to each sample
  df$Year <- rep.int(as.integer(year), length(df$Job.title))
  
  ## Cleaning the Names
  df$Name <- toupper(df$Name)
  comma_idxs <- grepl(pattern=",", x = df$Name)
  comma_names <- df$Name[comma_idxs]
  df$Name[comma_idxs] <- sapply(strsplit(comma_names, ", "), function(x) paste(x[2], x[1]))
  df$Name <- sapply(strsplit(as.character(df$Name), " "), function(x) paste(x[1], x[length(x)]))
  return(df)
}
```

```{r, echo=FALSE}
# Cleaning the data
data_2011 <- clean_data(data_2011, "2011")
data_2012 <- clean_data(data_2012, "2012")
data_2013 <- clean_data(data_2013, "2013")
data_2014 <- clean_data(data_2014, "2014")
data_2015 <- clean_data(data_2015, "2015")
data_2016 <- clean_data(data_2016, "2016")

# Combining the data into one dataset
total_data <- do.call(rbind, list(data_2011,data_2012,data_2013,data_2014,data_2015,data_2016))
```

#### Question 1: How has compensation changed over time for each position in appraisal?

To answer this question, the total pay and the total benefits will be grouped by job title and year.
Then, the data will be plotted independently and finally together to get a larger view of the changes.

```{r, echo=FALSE}
# adding total pay bins to the data

start <- min(total_data$Total.pay)
end <- as.integer(max(total_data$Total.pay) + 1)
splits <- as.integer(end / 4)

total_data$total.pay.bin <- cut(total_data$Total.pay, breaks=c(start, start + splits, start + 2*splits, start + 3*splits, end))

# adding total benefit bins to the data
start <- min(total_data$Total.benefits)
end <- as.integer(max(total_data$Total.benefits) + 1)
splits <- as.integer((end - start) / 4)

total_data$total.benefits.bin <- cut(total_data$Total.benefits, breaks=c(start, start + splits, start + 2*splits, start + 3*splits, end))
```

#### Analysis of Appraiser I

Appraiser I is the first position to which a trainee is promoted.  Here, and in each of the following
sections, total pay, total benefits, and promotion statistics will be explored.

##### Total Pay

```{r, echo=FALSE}
## Plotting boxplots for Appraiser I
ggplot(aes(x=factor(Year), y=Total.pay), data=subset(total_data, Job.title == "APPRAISER I")) +
  geom_boxplot() +
  ggtitle("Boxplots for Appraiser I Total Pay by Year")
```

```{r, echo=FALSE}
ggplot(aes(factor(Year), fill=total.pay.bin), data=subset(total_data, Job.title == "APPRAISER I")) +
  geom_bar() +
  ggtitle("Number of Appraiser I's by year")
```

###### Observations from Appraiser I Total Pay:

There is an odd dip from 2013 to 2014 in the total pay.  In 2014 the plot suggests
that many people with the largest salaries left for some reason (e.g. they 
were promoted, they quit, or they were laid off) and that many people were brought in 
with a lower salary than the average for the previous years.  While the distribution
for 2014 at the 3rd quartile extends to the first quartile of 2013, the mean is drawn
down to below $50,000.  The boxplot for 2015 suggest that the previous year's new recruits
got a raise or left because the minimum no longer reaches below $50,000. For 2016 it seems
like the entire population of Appraiser I employees were new recruits.

##### Total Benefits

```{r, echo=FALSE}
## Plotting boxplots for Appraiser I
ggplot(aes(x=factor(Year), y=Total.benefits), data=subset(total_data, Job.title == "APPRAISER I")) +
  geom_boxplot() +
  ggtitle("Boxplots for Appraiser I Total Benefits by Year")
```
```{r, echo=FALSE}
ggplot(aes(factor(Year), fill=total.benefits.bin), data=subset(total_data, Job.title == "APPRAISER I")) +
  geom_bar() +
  ggtitle("Number of Appraiser I's by Year for Benefits")
```

```{r, echo=FALSE}
########################################################
############  Helper Functions For Promotions ###########
#########################################################

# Function: set.intersection
# Returns: the intersection of two lists
# Fields: a = list, b = list

set.intersection <- function(a, b) {
  intersect <- vector()
   
  for (i in 1:length(a)) {
    if (a[i] %in% b) {
      intersect <- append(intersect, a[i])
    }
  }
  return(intersect)
}

# Function to find the distribution of the amount of time it
# took for employees to transition from one title to another

time.to.promotion.distribution <- function(position1, position2, df) {
  initial.position <- df[df$Job.title == position1,]
  promoted.position <- df[df$Job.title == position2,]
  promoted.people <- set.intersection(initial.position$Name, promoted.position$Name)
  
  years.to.promotion <- vector()

  for (i in 1:length(promoted.people)) {
    name <- promoted.people[i]
    promotion.year <- promoted.position$Year[promoted.position$Name == name][1]
    start.year <- initial.position$Year[initial.position$Name == name][1]
    years.to.promotion <- append(years.to.promotion, promotion.year - start.year)
  }
  return(years.to.promotion)
}

## Function that finds the number of people for each
## job

total.num.workers <- function(position, df) {
  num.workers <- df[df$Job.title == position,]
  return(length(num.workers$Name))
}

## Function that finds the number of people who get promoted
## from one position to another

num.workers.promoted <- function(position1, position2, df) {
  initial.position <- df[df$Job.title == position1,]
  promoted.position <- df[df$Job.title == position2,]
  promoted.people <- set.intersection(initial.position$Name, promoted.position$Name)
  return(length(promoted.people))
}

## Function: mean.years.to.promotion

summary.of.years.to.promotion <- function(distribution) {
  return(summary(distribution))
}
```

##### Examining Promotion Data For Appraiser Trainee's to Appraiser I

From the data we can see that there was at one time 25 Appraiser Trainees and
15 were promoted to Appraiser I (60% promotion rate).  As the
histogram shows below, for those people who were promoted, they took at most
one year to become an Appraiser I.

```{r, echo=FALSE}

trainer.to.appraiser <- time.to.promotion.distribution("APPRAISER TRAINEE", "APPRAISER I", total_data)
zero.to.one.df <- data.frame(trainer.to.appraiser)

total.num.workers("APPRAISER TRAINEE", total_data)
num.workers.promoted("APPRAISER TRAINEE", "APPRAISER I", total_data)
num.workers.promoted("APPRAISER TRAINEE", "APPRAISER I", total_data) / total.num.workers("APPRAISER TRAINEE", total_data)

ggplot(aes(x = trainer.to.appraiser), data=zero.to.one.df) +
  geom_histogram(binwidth = 0.5) +
  ggtitle("Histogram of Number of Years from Appraiser Trainee to Appraiser I") +
  xlab("Number of Years") +
  ylab("Number of People")
```

###### Observations from Total Benefits

The boxplots show that the benefits mirror the dipping effect in 2014 with the
steady rise afterward.

#### Analysis of Appraiser II

##### Total Pay
```{r, echo=FALSE}
## Plotting boxplots for Appraiser II
ggplot(aes(x=factor(Year), y=Total.pay), data=subset(total_data, Job.title == "APPRAISER II")) +
  geom_boxplot() +
  ggtitle("Boxplots for Appraiser II Total Pay by Year")
```

The outliers of this data make the trend in this data less apparent.  The next plot shows
the boxplots without outliers.

```{r, echo=FALSE}
## Plotting boxplots for Appraiser II
ggplot(aes(x=factor(Year), y=Total.pay), data=subset(total_data, Job.title == "APPRAISER II")) +
  geom_boxplot() +
  ylim(65000, 100000) +
  ggtitle("Boxplots for Appraiser II Total Pay by Year")
```

```{r, echo=FALSE}
ggplot(aes(factor(Year), fill=total.pay.bin), data=subset(total_data, Job.title == "APPRAISER II")) +
  geom_bar() +
  ggtitle("Number of Appraiser II's by Year for Pay")
```

##### Total Benefits

```{r, echo=FALSE}
ggplot(aes(x=factor(Year), y=Total.benefits), data=subset(total_data, Job.title == "APPRAISER II")) +
  geom_boxplot() +
  ggtitle("Boxplots for Appraiser II Total Benefits by Year")
```

```{r, echo=FALSE}
ggplot(aes(factor(Year), fill=total.benefits.bin), data=subset(total_data, Job.title == "APPRAISER II")) +
  geom_bar() +
  ggtitle("Number of Appraiser II's by Year for Benefits")
```

###### Observations from Appraiser II Plots:

The number of employees and their respective salaries are far less variable for 
Appraiser II's than Appraiser I's.  Again, the plots show a dip in mean salaries
in 2014.  The bar chart reveals that the exit one the employee that made close to
$100,000 had a significant effect on the change in the mean.  Overall, it appears
as if Total Pay for Appraiser II's is increasing.

##### Examining Promotion Data For Appraiser I to Appraiser II

From the data we can see that there was at one time 72 Appraiser I and
60 were promoted to Appraiser II (83.33% promotion rate).  As the
histogram shows below, for those people who were promoted, they took at most
one year to become an Appraiser I.  The mean number of years to promotion
is 2.93 and the median is 3.

```{r, echo=FALSE}
appraiser.1.to.appraiser.2 <- time.to.promotion.distribution("APPRAISER I", "APPRAISER II", total_data)
one.to.two.df <- data.frame(appraiser.1.to.appraiser.2)

total.num.workers("APPRAISER I", total_data)
num.workers.promoted("APPRAISER I", "APPRAISER II", total_data)
num.workers.promoted("APPRAISER I", "APPRAISER II", total_data) / total.num.workers("APPRAISER I", total_data)
summary.of.years.to.promotion(appraiser.1.to.appraiser.2)

ggplot(aes(x = appraiser.1.to.appraiser.2), data=one.to.two.df) +
  geom_histogram(binwidth = 0.5) +
  ggtitle("Histogram of Number of Years from Appraiser I to Appraiser II") +
  xlab("Number of Years") +
  ylab("Number of People")
```

#### Analysis of Appraiser III

##### Total Pay
```{r, echo=FALSE}
## Plotting boxplots for Appraiser II
ggplot(aes(x=factor(Year), y=Total.pay), data=subset(total_data, Job.title == "APPRAISER III")) +
  geom_boxplot() +
  ylim(65000, 100000) +
  ggtitle("Boxplots for Appraiser III Total Pay by Year")
```

```{r, echo=FALSE}
ggplot(aes(factor(Year), fill=total.pay.bin), data=subset(total_data, Job.title == "APPRAISER III")) +
  geom_bar() +
  ggtitle("Number of Appraiser III's by Year for Pay")
```

###### Observations for Total Pay:

Clearly more investigation needs to occur. In 2013, around one quarter of the 
Appraiser III's made close to $100,000, and by 2014 there are none.  We
start to see larger salaries pop up again in 2016 which suggests there will
be more next year.

##### Total Benefits

```{r, echo=FALSE}
## Plotting boxplots for Appraiser II
ggplot(aes(x=factor(Year), y=Total.benefits), data=subset(total_data, Job.title == "APPRAISER III")) +
  geom_boxplot() +
  ylim(20000, 50000) +
  ggtitle("Boxplots for Appraiser III Total Benefits by Year")
```
```{r, echo=FALSE}
ggplot(aes(factor(Year), fill=total.benefits.bin), data=subset(total_data, Job.title == "APPRAISER III")) +
  geom_bar() +
  ggtitle("Number of Appraiser III's by Year for Benefits")
```

##### Examining Promotion Data For Appraiser II to Appraiser III

From the data we can see that there was at one time 263 Appraiser II and
31 were promoted to Appraiser II (11.78% promotion rate).  The median
number of years to promotion was 4 and the mean was 3.29.

```{r, echo=FALSE}
appraiser.2.to.appraiser.3 <- time.to.promotion.distribution("APPRAISER II", "APPRAISER III", total_data)
appraiser.2.to.appraiser.3 <- appraiser.2.to.appraiser.3[appraiser.2.to.appraiser.3 > 0]
two.to.three.df <- data.frame(appraiser.2.to.appraiser.3)
colnames(two.to.three.df) <- c("num.years")

total.num.workers("APPRAISER II", total_data)
num.workers.promoted("APPRAISER II", "APPRAISER III", total_data)
num.workers.promoted("APPRAISER II", "APPRAISER III", total_data) / total.num.workers("APPRAISER II", total_data)
summary.of.years.to.promotion(appraiser.2.to.appraiser.3)

ggplot(aes(x = num.years), data=two.to.three.df) +
  geom_histogram(binwidth = 0.5) +
  ggtitle("Histogram of Number of Years from Appraiser II to Appraiser III") +
  xlab("Number of Years") +
  ylab("Number of People")
```



